Weak interspecific interactions in a sagebrush steppe: evidence from observations, models, and experiments

### 2017-01-30 removals/manuscript/removal_ms.tex
Final edits to ms and cover letter

\begin{enumerate}
\item A previous study of four dominant species from a sagebrush steppe community used historical data sets and multispecies population models to demonstrate weak interspecific interactions. The models predict little competitive release following species removals.
\item We conducted a removal experiment to test this prediction. We established new quadrats in the same location where the historical data were collected. We assigned half the new quadrats to a  perennial grass removal treatment and half to a treatment in which we removed the dominant shrub, \textit{Artemisia tripartita}. We modeled survival, growth and recruitment as functions of local neighborhood species composition along with an indicator variable for removal treatment. If our ``baseline" model, which accounts for local plant-plant interactions, explains responses to removal, then the fitted removal indicator effect should be non-signficant.
\item  For survival and recruitment, the removal treatment effects were never significantly positive, indicating that competitive release was not underestimated by the interspecific effects included in the baseline model. For  \textit{Poa secunda} recruitment, the removal treatment was actually negative, indicating that our baseline model overestimated competitive release.
\item For growth, the removal treatment effect was significant and positive for two species, \textit{Poa secunda} and \textit{Pseudoroegneria spicata}, indicating that the baseline model underestimated competitive release. However, including information about the location of these grass individuals with respect to removed \textit{A. tripartita} failed to improve the growth regressions, raising questions about the mechanisms driving the positive response to removal.
\item For three species, individual based models and integral projections models based on the vital rate regressions showed that removal treatment effects had little impact on population growth rates or simulated equilibrium cover following a species removal. For \textit{P. spicata}, the population models that included effects of \textit{A. tripartita} removal projected greater competitive release than the baseline models. 
\item \emph{Synthesis}: Although models based on observational data did not perfectly predict species responses to removal, the experimental results increase our confidence that competitive release is in fact small for at least three of our four study species. The implication is that interspecific resource competition is not the primary factor determining the dynamics of this community.

@@

To test whether our models, which are based on observational data, underestimate the strength of interspecific competition and competitive release, we conducted a removal experiment in a sagebrush steppe. We studied the responses of three perennial grass species to removal of sagebrush, and the response of sagebrush to removal of all grasses. We addressed two research questions: First, are the survival, growth or recruitment of our target species significantly different in control vs. removal treatments after accounting for differences in local species composition? Because our demographic models include the effects of neighborhood crowding, they should be able to predict how experimental removal of neighbors alters demographic performance. If the full effect of experimental removals is stronger than predicted, it means that our statistical models underestimate competitive release.  Our second question integrates across the full life cycle: how do short and long-term population-level responses to removals compare to projections from multispecies population models based on observational data? If models based on observational data adequately predict competitive release, the population dynamics they project should be similar to projections from models that include information from the experimental removal treatments.

@@

Formally, we modeled the survival probability, $S$, of genet $i$ in species $j$, group $g$, and removal treatment $h$, from time $t$ to $t+1$  as
\begin{equation}
\mbox{logit}(S_{ijgh,t+1}) = \gamma_{j,t}^S + \varphi_{jg}^S+  \chi_{jh}^S  + \beta_{j,t}^S u_{ij,t} +  
\left \langle \boldsymbol{\omega}_{j,t}^S, \boldsymbol{W}_{ij,t} \right \rangle 
\label{eqn:survReg}
\end{equation}
where $\gamma$ is a time-dependent intercept, $\varphi$ is the coefficient for the effect of spatial location group, $\chi$ is the removal treatment effect, $\beta$ is the coefficient that represents the effect of log genet size, $u$, on survival. $\boldsymbol{\omega}$ is a vector of interaction coefficients which determine the impact of crowding, $\boldsymbol{W}$, by each species on the focal species. The vector $\boldsymbol{W}$ includes crowding from the four dominant species,  \textit{A. tripartita}, \textit{H. comata}, \textit{P. spicata}, and \textit{P. secunda}, as well as two covariates representing total crowding from (1) all other perennial grasses and shrubs, which were mapped as polygons, and (2) all forb species, which were mapped as points. 
$\left \langle \boldsymbol{x, y} \right \rangle$ denotes the inner product of vectors $\boldsymbol{x}$ and $\boldsymbol{y}$, 
calculated as \texttt{sum(x*y)} in R.

Our growth model has a similar structure. The change in genet size from time $t$ to $t+1$ , conditional on survival, is given by:
\begin{equation}
u_{ijgh,t+1} = \gamma_{j,t}^G + \varphi_{jg}^G+  \chi_{jh}^G  + \beta_{j,t}^G u_{ij,t} + 
\left \langle  \boldsymbol{\omega}_{j,t}^G, \boldsymbol{W}_{ij,t} \right \rangle + \varepsilon_{ij,t}^G .
\label{eqn:growReg}
\end{equation}
To capture non-constant error variance in growth, we modeled the variance $\varepsilon$  about the growth curve (\ref{eqn:growReg})  as a nonlinear function of predicted genet size:
\begin{equation}
Var(\varepsilon_{ij,t}^G) = a \exp ^{(bu_{ij,t+1})} .
\label{eqn:growVar}
\end{equation}

We model the crowding experienced by a focal genet as a function of the distance to and size of neighbor genets. In previous work, we assumed that the decay of crowding with neighbor distance followed a Gaussian function \citep{chu_large_2015}, but here we use a data-driven approach \citep{teller_linking_2016}. We model the crowding experienced by genet $i$ of species $j$ from neighbors of species $m$ as the sum of neighbor areas across a set of concentric annuli, $k$, centered at the plant,
\begin{equation}
w_{ijm,k} = F_{jm}(d_{k})A_{i,k}     
\label{eqn:wik}
\end{equation}
where $F_{jm}$ is the competition kernel (described below) for effects of species $m$ on species $j$, 
$d_{k}$ is the average of the inner and outer radii of annulus $k$, 
and $A_{im,k}$ is the total area of genets of species $m$ in annulus $k$ around genet $i$. The total crowding on 
genet $i$ exerted by species $m$ is
\begin{equation}
W_{ijm}  =\sum_k {w_{ijm,k}} .
\label{eqn:wijm}
\end{equation} 
Note that $W_{ijj}$ gives intraspecific crowding. The $W$'s are then the components of the $\boldsymbol{W}$ vectors 
introduced as covariates in the survival (\ref{eqn:survReg}) and growth (\ref{eqn:growReg}) regressions.

We assume that competition kernels $F_{jm}(d)$ are non-negative and decreasing, so that distant plants have less effect 
than close plants. Otherwise, we let the data dictate the shape of the kernel by fitting a spline model 
using the methods of Teller et al. (2016). The shape of $F_{jm}$ is determined by a set of spline basis coefficients $\vec{b}_{jm}$
and a smoothing parameter $\eta$ that controls the complexity of the fitted kernel. 
Demographic models such as \eqref{eqn:survReg} then have $\gamma$, $\varphi$, $\chi$ , 
$\beta$, $\boldsymbol{\omega}$, $\vec{b}$ and $\eta$ as parameters to be fitted. We implemented this in the statistical computing environment, \texttt{R}, 
by making the spline coefficients and $\eta$ the arguments of an objective function that computes $\boldsymbol{W}$ using the input spline coefficients, 
calls the model-fitting functions \texttt{lmer} and/or \texttt{glmer} to fit the other parameters in the survival and growth regressions, 
and returns an approximate AIC value and model degrees of freedom ($df$) for survival and growth combined. We used the $\vec{b}$ values at the smoothest 
local minimum of AIC as a function of $df$, as in \cite{teller_linking_2016}. This approach assumes that one measure of crowding affects 
survival and growth. In addition, for fitting the kernels we assumed that survival and growth depended only on intraspecific crowding, and thus only fitted the
within-species competition kernels $F_{jj}$. Based on previous work \citep{adler_coexistence_2010}, we set all $F_{mj}$ equal to $F_{jj}$, meaning that 
the within-species competition kernel for species $j$ is also used to determine the effect of all other species on species $j$. We used data from all historical plots and contemporary control-treatment plots to estimate the competition kernels, which are shown in Fig. \ref{fig:CompKernels}.

Once we had estimated the competitions kernels, we used them to calculate the values of $\boldsymbol{W}$ for each individual, 
and fit the full survival and growth regressions, which include the interspecific interaction coefficients, $\boldsymbol{\omega}$. 
All genets in a quadrat were included in calculating $W$, but plants located within 5 cm of quadrat edges were not used in fitting. 
We fit the models using the \texttt{R-INLA} package for R \citep{rue_approximate_2009}. We used the same approach to explore variations on the growth regressions, such as the addition of a year-by-treatment interaction or information about the locations of individual grasses relative to removed \textit{A. tripartita}. We also compared the species composition of the control and removal plots at both the quadrat and neighborhood ($\boldsymbol{W}$) scales.

We model recruitment at the quadrat level rather than at the individual genet level because the mapped data do not allow us to determine which recruits were produced by which potential parent genets. We assume that the number of individuals, $y$, of species $j$ recruiting at time $t+1$ in the location $q$ follows a negative binomial distribution:
\begin{equation}
y_{jq,t+1}= NegBin(\lambda_{jq,t+1},\theta) 	   
\label{eqn:recrDataModel}
\end{equation}
where $\lambda$ is the mean intensity and $\theta$ is the size parameter. In turn, $\lambda$ depends on the composition of the quadrat in the previous year:
\begin{equation}
\lambda_{jq,t+1} = C'_{jq,t} \exp{\left(\gamma_{j,t}^R +  \varphi_{jg}^R + \chi_{jm}^R + 
\left \langle \boldsymbol{\omega}^R , \boldsymbol{\sqrt{C'}_{q,t}} \right \rangle \right) }
\label{eqn:recrProcessModel}
\end{equation}
where the superscript $R$ refers to Recruitment, $C'_{jq,t}$ is the `effective cover' (cm$^2$) of species $j$ in quadrat $q$ at time $t$, $\gamma$ is a time-dependent intercept, $\varphi$ is a coefficient for the effect of spatial location,
$\chi$ is the removal treatment effect, $\boldsymbol{\omega}$ is a vector of coefficients that determine the strength of intra- and interspecific density-dependence, and $\boldsymbol{C'}$ is the vector of ``effective'' cover of each species in the community. Eqn (\ref{eqn:recrProcessModel}) is essentially the Ricker model for discrete time population growth \citep{ricker_stock_1954}, except that we use the square root of cover in the density-dependent term. We previously used a Ricker equation \citep{adler_coexistence_2010}, but using the square-root transformation instead gave a lower Deviance Information Criterion (DIC) value \citep{spiegelhalter_bayesian_2002} for the recruitment model. Following previous work \citep{adler_coexistence_2010}, we treated year and spatial group variables as random factors, allowing intercepts to vary among years and spatial locations.

@@

The vital rate regressions provide a qualitative test of removal treatment effects: Does removal have an effect on a vital rate that goes above and beyond the effects of species interactions that are explicit in the model? However, the population-level response by each species integrates the effects of the removal treatment on all three vital rates. We used an individual-based model (IBM) to compare observed and predicted changes 
in population size (cover) from one year to the next.

To simulate changes in cover in each quadrat from year $t$ to year $t+1$, we initialized the IBM with the actual configuration of the four focal species (genet sizes and locations) observed in year $t$. We then projected the model forward one year, applying the random year effects corresponding to year $t$ and the spatial random effects corresponding to the quadrat of interest. Because we were interested in comparing model predictions to observations, and were not interested in the effects of demographic stochasticity, we used a deterministic version of the model (e.g., recruitment is the $\lambda$ of (\ref{eqn:recrProcessModel}), rather than a random draw from a Poisson distribution with a mean of $\lambda$).
We first applied the recruitment regression to determine the expected number of new recruits at time $t+1$, which we then multiplied by the mean size of recruits observed for the focal species to get expected cover of new recruits. We then used the survival regression to predict the survival probability of each genet, and the growth regression to predict the change in size of each genet. Expected cover at $t+1$ is expected area of new recruits, plus the sum of the survival probabilities multiplied by the expected sizes. Because we do not carry these simulations forward beyond one time step, we do not need to assign locations to the new recruits.

@@

To quantify the long-term, population-level consequences of the removal effects, we built and then simulated an integral projection model (IPM). In our IPM, the population of species $j$ is represented by a density function $n(u_{j,t})$ which gives the density of genets of size $u$ at time $t$, with genet size on a natural-log scale, i.e. $n(u_{j,t})du$ is the number of genets whose log area is between $u$ and $u+du$, hence their area 
(on arithmetic scale) is between $exp(u_j)$ and $exp(u_j+du)$. The size distribution function at time $t+1$ is given by
\begin{equation}
n(v_{j,t+1})=\int_{L_j}^{U_j} k_j (v_j,u_j,{\vec{n}})n(u_{j,t})   
\label{eqn:IPM}
\end{equation}
where the kernel $k_j (v_j,u_j,\vec{n}_j)$ describes all possible transitions from size $u$ to $v$ and $\vec{n}$ 
is the set of size-distribution functions for all species in the community. The integral is evaluated over all possible sizes from a 
lower size limit $L$ to an upper size limit $U$ that extends past the range of observed sizes.

The kernel is constructed from the fitted survival ($S$), growth ($G$), and recruitment ($R$) models, 
\begin{equation}
k_j(v_j,u_j,\vec{n})=S_j(u_j,\vec{n})G_j(v_j,u_j,\vec{n})+R_j(v_j,u_j,\vec{n}) du_j
\label{eqn:IPM} 
\end{equation}
$S$ is given by (Eqn. \ref{eqn:survReg}) and $G$ is given by (Eqn. \ref{eqn:growReg}), using an expected neighborhood crowding
index calculated from the size distributions of the competing species. In the vital rate regressions and in the IBM, each 
individual has a unique neighborhood crowding index based on the spatial locations and sizes of neighboring plants. This 
measure cannot be extended to the IPM, which tracks individual size but not location. Instead we used a 
mean-field approximation that captures the essential features of spatially-explicit neighborhood interactions \citep{adler_coexistence_2010}. 
We found that in both the observed data and IBM simulations, heterospecific individuals were approximately randomly distributed with respect to each other, 
but conspecific individuals displayed a non-random, size-dependent patterns: small genets were randomly distributed, while large genets 
were segregated from each other. The overdispersion of large conspecific genets was introduced into the IPM 
with a `no-overlap' rule \citep{adler_coexistence_2010}. In \ref{suppMethods} we explain how the spline competition kernels with the no-overlap rule kernels were used to compute neigbhorhood crowding index as a function of individual size for the IPM.

For recruitment in the IPM, the factor $ \Phi_j  = \exp{\left(\gamma_{j,t}^R +  \varphi_{jg}^R + \chi_{jm}^R + 
\left \langle \boldsymbol{\omega}^R , \boldsymbol{\sqrt{C'_{q,t}}} \right \rangle \right) }$
in eqn. (\ref{eqn:recrDataModel}) gives the expected 
area of new recruits per quadrat, per unit area of parents. To incorporate this recruitment function into the IPM, we assumed that individual fecundity increases linearly with individual area (equal to $e^u$ for a size-$u$ individual). The recruitment kernel is thus
$R_j(v_j,u_j,\vec{n})=c_{0,j}(v_j)e^{u_j}\Phi_j$ \citep{adler_coexistence_2010} where $c_{0,j}$ is the
size distribution of new recruits. $\Phi_j$ is calculated from eqn. (\ref{eqn:recrDataModel}) with total cover values computed from the population densities as $C'_{j,t} = \int e^u n_j(u,t) du$.

We simulated the IPM to estimate equilibrium abundances under three different scenarios. For all IPM simulations, we set the spatial random effects to zero, meaning that we are simulating dynamics on a hypothetical, average site. The first scenario featured all four species and our ``baseline'' model for control plots (no removal effects). We initialized all species with arbitrarily low cover and ran the model for 500 time steps (years) to allow the community to reach an equilibrium. To incorporate temporal environmental variation, at each time step we randomly selected one of the 30 sets of random year effects for the survival, growth and recruitment models (``kernel selection'' in the terminology of \citealt{metcalf_statistical_2015}). We calculated the equilibrium cover for each species by simulating an additional 2000 time steps and averaging each species' cover over this period. For the second scenario, we set the cover of \textit{A. tripartita} to zero and then repeated the procedure. This simulation estimates the competitive release of grasses following \textit{A. tripartita} removal according to the baseline model (no removal effects). For the third scenario, we again set  \textit{A. triparitia} to zero, but this time we projected the model using the vital rate regressions that include removal effects. If our baseline model successfully predicts competitive release observed in the removal experiment, then the results of the second and third scenarios should be similar. To the extent that the baseline model underestimates (or overestimates) the actual competitive release, the results of the second and third scenarios will differ. As with the IBM, we repeated these simulations using maximum removal treatment effects represented by parameters values at the 5\% or 95\% confidence limit farthest from zero. We did not simulate long-term \textit{A. tripartita} response to grass removal because the vital rate regressions for \textit{A. tripartita} showed that the baseline model effectively predicted removal effects.

@@

Trends in cover by species and removal treatment were variable. For example, grass removal had little effect on the cover of \textit{A. tripartita} 
(Fig. \ref{fig:CoverTrends}), but shrub removal did appear to cause an increase in the cover of \textit{P. spicata}. 
We tested the effects of removal on cover by analyzing year-to-year changes in (log) cover, a measure of population growth rate (Fig. \ref{fig:CoverChange}). The removal treatments had no significant effect on 
\textit{A. tripartita} (95\% confidence interval on removal treatment effect: -0.75, 0.56) or 
\textit{P. secunda} (95\% CI: -0.41, 0.25) population growth, but did increase 
\textit{H. comata} (95\% CI: 0.01, 0.65) and
 \textit{P. spicata} (95\% CI: 0.01, 0.40) population growth 
(Table \ref{table:coefficients}). However, these treatment comparisons do not address our questions about whether responses to removal were 
adequately predicted by models fit to observational data on non-removal plots.

@@

PBA and ARK designed the experiment and supervised data collection with the help of JBT, PBA, SPE, GH and BT analyzed data, PBA wrote the first draft of the manuscript and all authors contributed to editing.

>>>>

\begin{enumerate}
\item A previous study of four dominant species from a sagebrush steppe community used historical data sets and multispecies population models to demonstrate weak interspecific interactions. The models predict little competitive release following species removals.
\item We conducted a removal experiment to test this prediction. We established new quadrats in the same location where the historical data were collected. We assigned half the new quadrats to a  perennial grass removal treatment and half to a treatment in which we removed the dominant shrub, \textit{Artemisia tripartita}. We modeled survival, growth and recruitment as functions of local neighborhood species composition along with an indicator variable for removal treatment. If our ``baseline" model, which accounts for local plant-plant interactions, explains responses to removal, then the fitted removal indicator effect should be non-signficant.
\item  For survival and recruitment, the removal treatment effects were never significantly positive, indicating that competitive release was not underestimated by the interspecific effects included in the baseline model. For  \textit{Poa secunda} recruitment, the removal treatment was actually negative, indicating that our baseline model overestimated competitive release.
\item For growth, the removal treatment effect was significant and positive for two species, \textit{Poa secunda} and \textit{Pseudoroegneria spicata}, indicating that the baseline model underestimated competitive release. However, including information about the location of these grass individuals with respect to removed \textit{A. tripartita} failed to improve the growth regressions, raising questions about the mechanisms driving the positive response to removal.
\item For three species, individual based models and integral projections models based on the vital rate regressions showed that removal treatment effects had little impact on population growth rates or simulated equilibrium cover following a species removal. For \textit{P. spicata}, the population models that included effects of \textit{A. tripartita} removal projected greater competitive release than the baseline models. 
\item \emph{Synthesis}: Although models based on observational data did not perfectly predict species responses to removal, the experimental results increase our confidence that competitive release is in fact small for at least three of our four study species. The implication is that interspecific resource competition is not a primary factor determining the dynamics of this community. The success of the models also supports the use of multispecies models, based on long-term observational studies, to characterize competitive interactions in a community.

@@

To test whether our models, which are based on observational data, underestimate the strength of interspecific competition and competitive release, we conducted a removal experiment in a sagebrush steppe. We studied the responses of three perennial grass species to removal of sagebrush, and the response of sagebrush to removal of all grasses. We addressed two research questions: First, are the survival, growth or recruitment of our target species significantly different in control vs. removal treatments after accounting for differences in local species composition? Because our demographic models include the effects of neighborhood crowding, they should be able to predict how experimental removal of neighbors alters demographic performance. If the full effect of experimental removals is stronger than predicted, it means that our statistical models underestimate competitive release.  Our second question integrates across the full life cycle: how do short and long-term population-level responses to removals compare to projections from multispecies population models based on observational data? If models based on observational data adequately predict competitive release, the population dynamics they project should be similar to projections from models that include information from the experimental removal treatments, a result which would validate the use of models and long-term data sets to study interspecific interactions.

@@

Formally, we modeled the survival probability, $S$, of genet $i$ in species $j$, group $g$, and removal treatment $h$, from time $t$ to $t+1$  as
\begin{equation}
\mbox{logit}(S_{ijgh,t+1}) = \gamma_{j,t}^S + \varphi_{jg}^S+  \chi_{jh}^S  + \beta_{j,t}^S u_{ij,t} +  
% \left \langle \boldsymbol{\omega}_{j,t}^S, \boldsymbol{W}_{ij,t} \right \rangle 
\sum \limits_{j} \omega_{j}^S {W}_{ij,t}
\label{eqn:survReg}
\end{equation}
where $\gamma$ is a time-dependent intercept, $\varphi$ is the coefficient for the effect of spatial location group, $\chi$ is the removal treatment effect, $\beta$ is the coefficient that represents the effect of log genet size, $u$, on survival. 
%$\boldsymbol{\omega}$ is a vector of interaction coefficients which determine the impact of crowding, $\boldsymbol{W}$, by each species on the focal species. 
$\omega_j$ is an interaction coefficient which determines the impact of crowding, $W_j$, by species $j$ on the focal species. 
Our model includes crowding from the four dominant species,  \textit{A. tripartita}, \textit{H. comata}, \textit{P. spicata}, and \textit{P. secunda}, as well as two covariates representing total crowding from (1) all other perennial grasses and shrubs, which were mapped as polygons, and (2) all forb species, which were mapped as points. 
%$\left \langle \boldsymbol{x, y} \right \rangle$ denotes the inner product of vectors $\boldsymbol{x}$ and $\boldsymbol{y}$, 
%calculated as \texttt{sum(x*y)} in R.

Our growth model has a similar structure. The change in genet size from time $t$ to $t+1$ , conditional on survival, is given by:
\begin{equation}
u_{ijgh,t+1} = \gamma_{j,t}^G + \varphi_{jg}^G+  \chi_{jh}^G  + \beta_{j,t}^G u_{ij,t} + 
%\left \langle  \boldsymbol{\omega}_{j,t}^G, \boldsymbol{W}_{ij,t} \right \rangle + \varepsilon_{ij,t}^G .
\sum \limits_{j} \omega_{j}^G {W}_{ij,t} + \varepsilon_{ij,t}^G .
\label{eqn:growReg}
\end{equation}
To capture non-constant error variance in growth, we modeled the variance $\varepsilon$  about the growth curve (\ref{eqn:growReg})  as a nonlinear function of predicted genet size:
\begin{equation}
Var(\varepsilon_{ij,t}^G) = a \exp ^{(bu_{ij,t+1})} .
\label{eqn:growVar}
\end{equation}

We model the crowding experienced by a focal genet as a function of the distance to and size of neighbor genets. In previous work, we assumed that the decay of crowding with neighbor distance followed a Gaussian function \citep{chu_large_2015}, but here we use a data-driven approach \citep{teller_linking_2016}. We model the crowding experienced by genet $i$ of species $j$ from neighbors of species $m$ as the sum of neighbor areas across a set of concentric annuli, $k$, centered at the plant,
\begin{equation}
w_{ijm,k} = F_{jm}(d_{k})A_{i,k}     
\label{eqn:wik}
\end{equation}
where $F_{jm}$ is the competition kernel (described below) for effects of species $m$ on species $j$, 
$d_{k}$ is the average of the inner and outer radii of annulus $k$, 
and $A_{im,k}$ is the total area of genets of species $m$ in annulus $k$ around genet $i$. The total crowding on 
genet $i$ exerted by species $m$ is
\begin{equation}
W_{ijm}  =\sum_k {w_{ijm,k}} .
\label{eqn:wijm}
\end{equation} 
Note that $W_{ijj}$ gives intraspecific crowding. The $W$'s are then % the components of the $\boldsymbol{W}$ vectors 
introduced as covariates in the survival (\ref{eqn:survReg}) and growth (\ref{eqn:growReg}) regressions.

We assume that competition kernels $F_{jm}(d)$ are non-negative and decreasing, so that distant plants have less effect 
than close plants. Otherwise, we let the data dictate the shape of the kernel by fitting a spline model 
using the methods of Teller et al. (2016). The shape of $F_{jm}$ is determined by a set of spline basis coefficients $\vec{b}_{jm}$
and a smoothing parameter $\eta$ that controls the complexity of the fitted kernel. 
Demographic models such as \eqref{eqn:survReg} then have $\gamma$, $\varphi$, $\chi$ , 
$\beta$, $\boldsymbol{\omega}$, $\vec{b}$ and $\eta$ as parameters to be fitted. We implemented this in the statistical computing environment, \texttt{R}, 
by making the spline coefficients and $\eta$ the arguments of an objective function that computes ${W}$ using the input spline coefficients, 
calls the model-fitting functions \texttt{lmer} and/or \texttt{glmer} to fit the other parameters in the survival and growth regressions, 
and returns an approximate AIC value and model degrees of freedom ($df$) for survival and growth combined. We used the $\vec{b}$ values at the smoothest 
local minimum of AIC as a function of $df$, as in \cite{teller_linking_2016}. This approach assumes that one measure of crowding affects 
survival and growth. In addition, for fitting the kernels we assumed that survival and growth depended only on intraspecific crowding, and thus only fitted the
within-species competition kernels $F_{jj}$. Based on previous work with parametric kernels, \citep{adler_coexistence_2010}, we know that interspecific effects are rarely significantly different from zero, and there is not enough signal to fit interspecific kernels nonparametrically. Therefore, we set all $F_{mj}$ equal to $F_{jj}$, meaning that 
the within-species competition kernel for species $j$ is also used to determine the effect of species $j$ on all other species. We used data from all historical plots and contemporary control-treatment plots to estimate the competition kernels, which are shown in Fig. \ref{fig:CompKernels}.

Once we had estimated the competitions kernels, we used them to calculate the values of all ${W_j}$ for each individual, 
and fitted the full survival and growth regressions, which include the interspecific interaction coefficients, $\omega_j$. 
All genets in a quadrat were included in calculating the $W$'s, but plants located within 5 cm of quadrat edges were not used in fitting. 
We fit the models using the \texttt{R-INLA} package for R \citep{rue_approximate_2009}. We used the same approach wen exploring variations on the growth regressions, such as the addition of a year-by-treatment interaction or information about the locations of individual grasses relative to removed \textit{A. tripartita}. We also compared the species composition of the control and removal plots at both the quadrat and neighborhood ($W$) scales.

We model recruitment at the quadrat level rather than at the individual genet level because the mapped data do not allow us to determine which recruits were produced by which potential parent genets. We assume that the number of individuals, $y$, of species $j$ recruiting at time $t+1$ in the location $q$ follows a negative binomial distribution:
\begin{equation}
y_{jq,t+1}= NegBin(\lambda_{jq,t+1},\theta) 	   
\label{eqn:recrDataModel}
\end{equation}
where $\lambda$ is the mean intensity and $\theta$ is the size parameter. In turn, $\lambda$ depends on the composition of the quadrat in the previous year:
\begin{equation}
\lambda_{jq,t+1} = C'_{jq,t} \exp{\left(\gamma_{j,t}^R +  \varphi_{jg}^R + \chi_{jm}^R + 
%\left \langle \boldsymbol{\omega}^R , \boldsymbol{\sqrt{C'}_{q,t}} \right \rangle \right) }
\sum \limits_{j} \omega_{j}^R \sqrt{C'}_{ij,t} \right)}
\label{eqn:recrProcessModel}
\end{equation}
where the superscript $R$ refers to Recruitment, $C'_{jq,t}$ is the `effective cover' (cm$^2$) of species $j$ in quadrat $q$ at time $t$, $\gamma$ is a time-dependent intercept, $\varphi$ is a coefficient for the effect of spatial location,
$\chi$ is the removal treatment effect, $\omega_j$ is the coefficient that determines the density-dependent effect of species $j$ on the focal species, and $C'_j$ is the ``effective'' cover of species $j$ in the quadrat. Eqn (\ref{eqn:recrProcessModel}) is essentially the Ricker model for discrete time population growth \citep{ricker_stock_1954}, except that we use the square root of cover in the density-dependent term. We previously used a Ricker equation \citep{adler_coexistence_2010}, but using the square-root transformation instead gave a lower Deviance Information Criterion (DIC) value \citep{spiegelhalter_bayesian_2002} for the recruitment model. Following previous work \citep{adler_coexistence_2010}, we treated year and spatial group variables as random factors, allowing intercepts to vary among years and spatial locations.

@@

The vital rate regressions provide a qualitative test of removal treatment effects: Does removal have an effect on a vital rate that goes above and beyond the effects of species interactions that are explicit in the model? However, the population-level effects of removal integrate the effects of the removal treatment on all three vital rates. We used an individual-based model (IBM) to predict changes in population size (cover) from one year to the next in removal and control quadrats, and compared these predictions with observed changes.

To simulate changes in cover in each quadrat from year $t$ to year $t+1$, we initialized the IBM with the actual configuration of the four focal species (genet sizes and locations) observed in year $t$. We then projected the model forward one year, applying the random year effects corresponding to year $t$ and the spatial random effects corresponding to the quadrat of interest. Because the main goal of this comparison is to look for systematic deviations between model predictions and observations, we used a deterministic version of the model (e.g., recruitment is the $\lambda$ of (\ref{eqn:recrProcessModel}), rather than a random draw from a Poisson distribution with a mean of $\lambda$).
We first applied the recruitment regression to determine the expected number of new recruits at time $t+1$, which we then multiplied by the mean size of recruits observed for the focal species to get expected cover of new recruits. We then used the survival regression to predict the survival probability of each genet, and the growth regression to predict the change in size of each genet. Expected cover at $t+1$ is expected area of new recruits, plus the sum of the survival probabilities multiplied by the expected sizes. Because we do not carry these simulations forward beyond one time step, we do not need to assign locations to the new recruits.

@@

To quantify the long-term, population-level consequences of the removal effects, we built and then simulated an integral projection model (IPM). In our IPM, the population of species $j$ is represented by a density function $n(u_{j,t})$ which gives the density of genets of size $u$ at time $t$, with genet size on a natural-log scale, i.e. $n(u_{j,t})du$ is the number of genets whose log area is between $u$ and $u+du$, hence their area 
(on arithmetic scale) is between $exp(u_j)$ and $exp(u_j+du)$. The size distribution function at time $t+1$ is given by
\begin{equation}
n(v_{j,t+1})=\int_{L_j}^{U_j} k_j (v_j,u_j,{\vec{n}})n(u_{j,t})du_j   
\label{eqn:IPM}
\end{equation}
where the kernel $k_j (v_j,u_j,\vec{n}_j)$ describes all possible transitions from size $u$ to $v$ and $\vec{n}$ 
is the set of size-distribution functions for all species in the community. The integral is evaluated over all possible sizes from a 
lower size limit $L$ to an upper size limit $U$ that extends past the range of observed sizes.

The kernel is constructed from the fitted survival ($S$), growth ($G$), and recruitment ($R$) models, 
\begin{equation}
k_j(v_j,u_j,\vec{n})=S_j(u_j,\vec{n})G_j(v_j,u_j,\vec{n})+R_j(v_j,u_j,\vec{n}) 
\label{eqn:IPM} 
\end{equation}
$S$ is given by (Eqn. \ref{eqn:survReg}) and $G$ is given by (Eqn. \ref{eqn:growReg}), using an expected neighborhood crowding
index calculated from the size distributions of the competing species. In the vital rate regressions and in the IBM, each 
individual has a unique neighborhood crowding index based on the spatial locations and sizes of neighboring plants. This 
measure cannot be extended to the IPM, which tracks individual size but not location. Instead we used a 
mean-field approximation that captures the essential features of spatially-explicit neighborhood interactions \citep{adler_coexistence_2010}. 
We found that in both the observed data and IBM simulations, heterospecific individuals were approximately randomly distributed with respect to each other, 
but conspecific individuals displayed a non-random, size-dependent patterns: small genets were randomly distributed, while large genets 
were segregated from each other. The overdispersion of large conspecific genets was introduced into the IPM 
with a `no-overlap' rule \citep{adler_coexistence_2010}. In \ref{suppMethods} we explain how the spline competition kernels with the no-overlap rule were used to compute neighborhood crowding index as a function of individual size for the IPM.

For recruitment in the IPM, the factor $ \Phi_j  = \exp{\left(\gamma_{j,t}^R +  \varphi_{jg}^R + \chi_{jm}^R + 
\left \langle \boldsymbol{\omega}^R , \boldsymbol{\sqrt{C'_{q,t}}} \right \rangle \right) }$
in eqn. (\ref{eqn:recrDataModel}) gives the expected 
area of new recruits per quadrat, per unit area of parents. To incorporate this recruitment function into the IPM, we assumed that individual fecundity increases linearly with individual area (equal to $e^u$ for a size-$u$ individual). The recruitment kernel is thus
$R_j(v_j,u_j,\vec{n})=c_{0,j}(v_j)e^{u_j}\Phi_j$ \citep{adler_coexistence_2010} where $c_{0,j}$ is the
size distribution of new recruits. $\Phi_j$ is calculated from eqn. (\ref{eqn:recrDataModel}) with total cover values computed from the population densities as $C'_{j,t} = \int e^u n(u_{j,t}) du$.

We simulated the IPM to estimate equilibrium abundances under three different scenarios. For all IPM simulations, we set the spatial random effects to zero, meaning that we are simulating dynamics on a hypothetical, average site. The first scenario featured all four species and our ``baseline'' model for control plots (no removal effects). We initialized all species with arbitrarily low cover and ran the model for 500 time steps (years) to allow the community to reach an equilibrium. To incorporate temporal environmental variation, at each time step we randomly selected one of the 30 sets of random year effects for the survival, growth and recruitment models (``kernel selection'' in the terminology of \citealt{metcalf_statistical_2015}). We calculated the equilibrium cover for each species by simulating an additional 2000 time steps and averaging each species' cover over this period. For the second scenario, we set the cover of \textit{A. tripartita} to zero and then repeated the procedure. This simulation estimates the competitive release of grasses following \textit{A. tripartita} removal according to the baseline model (no removal effects). For the third scenario, we again set  \textit{A. triparitia} cover to zero, but this time we projected the model using the vital rate regressions that include removal effects. If our baseline model successfully predicts competitive release observed in the removal experiment, then the results of the second and third scenarios should be similar. To the extent that the baseline model underestimates (or overestimates) the actual competitive release, the results of the second and third scenarios will differ. As with the IBM, we repeated these simulations using maximum removal treatment effects represented by parameters values at the 5\% or 95\% confidence limit farthest from zero. We did not simulate long-term \textit{A. tripartita} response to grass removal because the vital rate regressions for \textit{A. tripartita} showed that the baseline model effectively predicted removal effects.

@@

Trends in cover by species and removal treatment were variable. For example, grass removal had little effect on the cover of \textit{A. tripartita} 
(Fig. \ref{fig:CoverTrends}), but shrub removal did appear to cause an increase in the cover of \textit{P. spicata}. 
We tested the effects of removal on cover by analyzing year-to-year changes in (log) cover, a measure of population growth rate (Fig. \ref{fig:CoverChange}). The removal treatments had no significant effect on 
\textit{A. tripartita} (95\% confidence interval on removal treatment effect: -0.75, 0.56) or 
\textit{P. secunda} (95\% CI: -0.41, 0.25) population growth, but did increase 
\textit{H. comata} (95\% CI: 0.01, 0.65) and
 \textit{P. spicata} (95\% CI: 0.01, 0.40) population growth 
(Table \ref{table:coefficients}). However, these treatment comparisons do not address our questions about whether responses to removal were 
adequately predicted by models fit to historical, observational data.

@@

PBA and ARK designed the experiment and supervised data collection with the help of JBT; PBA, SPE, GH and BT analyzed data; PBA wrote the first draft of the manuscript and all authors contributed to editing.
